---
title: "MA5112_RNA.rmd"
author: "Barry"
date: "09/01/2020"
output: html_document
---

```{R}
library(plyr)
library(biomaRt)
library(tximport)
library(rhdf5)
library(gplots)
library(DESeq2)
library(igraph)
library(RColorBrewer)
library(limma)
library(WGCNA)
library(IHW)
library(PCAtools)
library(tidyverse)
library(org.Hs.eg.db)
library(devtools)
library(pheatmap)
#install_github("ctlab/fgsea") #not on bioconductor or R repo
library(fgsea)
library(GOplot)
library(pvclust)
```
# Read in Kallisto Abundance Files
```{r}
dir <- ("~/Kallisto_Quantification")
samples <- read.table(file.path(dir, "sra_runtable.csv"), 
                        sep=",", header=T, row.names="Sample")
files <- file.path(dir, rownames(samples), "abundance.h5")
names(files) <- paste0(rownames(samples))
```
# Create mart object
# Use Mart object to retrive the ENSEMBL Gene IDS corresponding to the ENSEMBL Transcript IDs
```{r}
mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")
results <- getBM(attributes = c("ensembl_transcript_id_version", "ensembl_gene_id"), mart = mart)
```
# The tximport package converts transcript abundance to gene counts
```{r}
tx2gene <- results[, 1:2]
#TXImport converts transcript quantification to gene level counts
txi <- tximport(files, type = "kallisto", tx2gene = tx2gene)
```
# Create the DESeq2 'dds' object for analysis
# Model tests condition while controlling for Rep
# Convert numericals to factors
```{r}
#Create the dds object for DESeq2 analysis
samples$Rep <- as.factor(samples$Rep)
dds <- DESeqDataSetFromTximport(txi, colData = samples, design = ~ Rep + Condition )
dds$type <- relevel(dds$Condition, ref = "control") 
keep <- rowSums(counts(dds)) >= 20 #remove rows with counts less than 20
dds <- dds[keep,]
dds <- DESeq(dds) 

rm(txi)

pdf(file="~/Plots/Dispersion.pdf")
plotDispEsts(dds)
dev.off()
```
# Extract the normalised counts from the dds object and log2 the data to do a quick QC
```{r}
cts <- counts(dds, normalized=T)
log_counts <- log2(cts + 1)
```
# Store the counts. Handy to so we dont have to Run DESeq2 every time (can be computationally expensive on large datsets).
```{r}
cts_out <- as.data.frame(cts)
cts_out <- tibble::rownames_to_column(cts_out, "ensembl_gene_id")
write.table(cts_out, "~/Results/norm_counts.txt", sep="\t", quote = F, row.names = F)
```
# Store the counts. Handy to so we dont have to Run DESeq2 every time (can be computationally expensive on large datsets).
```{r}
cts_out <- as.data.frame(cts)
cts_out <- tibble::rownames_to_column(cts_out, "ensembl_gene_id")
write.table(cts_out, "~/Results/norm_counts.txt", sep="\t", quote = F, row.names = F)
```
# Sample to Sample correlation heatmap
```{r}
num_conditions <- nlevels(samples$Condition)
pal <- colorRampPalette(brewer.pal(num_conditions, "Set1"))(num_conditions)
cond_colors <- pal[as.integer(samples$Condition)]
pdf(file="~/Plots/sample_heatmap.pdf")
heatmap.2(cor(log_counts), RowSideColors=cond_colors,
          trace='none', main='Sample correlations (log2-transformed)')
dev.off()
```
# sample hierarchical clustering (correlation between clusters formed)
```{R}
result <- pvclust(log_counts, method.dist = "cor", method.hclust = "average", nboot = 500)
pdf(file="~/Plots/sample_hclust.pdf")
plot(result)
pvrect(result)
dev.off()
```
# Run PCA on the samples. Which samples do we expect to have the best results.
```{r}
p <- pca(log_counts, metadata = samples)
pdf(file="~/Plots/PCA.pdf")
biplot(p,
  colby = 'Condition', colkey = c('melanoma'='royalblue', 'control'='red1', 'lung'='forestgreen'),
  hline = 0, vline = 0,
  legendPosition = 'right', legendLabSize = 12, legendIconSize = 8.0,
  drawConnectors = TRUE,
  title = 'PCA bi-plot',
  subtitle = 'PC1 versus PC2')
dev.off()
```
# Set up the DESeq2 results object. Specify the contrast you want to make
```{r}
lung_v_ctrl <- results(dds, filterFun=ihw, alpha=0.05, c("Condition", "lung", "control"))
summary(lung_v_ctrl)
melanoma_v_ctrl <- results(dds, filterFun=ihw, alpha=0.05, c("Condition", "melanoma", "control"))
summary(melanoma_v_ctrl)
dds$condition <- relevel(dds$Condition, ref = "melanoma")
lung_v_melanoma <- results(dds, filterFun=ihw, alpha=0.05, c("Condition", "lung", "melanoma"))
summary(lung_v_melanoma)
```
# insect p-value histogram (distribution of null hypothesis)
```{r}
#pvalue histogram
pdf(file="~/Results/p-val_hist.pdf")
hist(lung_v_ctrl$pvalue, breaks=50, col="seagreen", xlab="P-Value (Fold change) A549 vs ctrl", main="Distribution of P-Values") 
abline(v=c(0.05),col="black",lwd=2,lty=2)
legend("topright", "P-Value <0.05",lwd=2,lty=2)
dev.off()
```
# Volcano plot for Visualization of adj-pvalue / Log2FC.
```{r}
#volcano before remove objects
resOrdered <- as.data.frame(lung_v_ctrl[order(lung_v_ctrl$padj),])
pdf(file="~/Results/A549_v_Ctrl_volcano.pdf")
plot(resOrdered$log2FoldChange, resOrdered$padj, col="black", pch=1, cex=1.1,  main="A549 vs Ctrl Volcano Plot")
fsig <-rownames(subset(resOrdered, log2FoldChange > 1.5 | log2FoldChange < -1.5,))
fsig_resOrdered_plot=resOrdered[fsig,]
points(fsig_resOrdered_plot$log2FoldChange, fsig_resOrdered_plot$padj, col="green", pch=19, cex=1.1)
psig <-rownames(subset(resOrdered, pvalue < 0.05))
psig_resOrdered_plot=resOrdered[psig,]
points(psig_resOrdered_plot$log2FoldChange, psig_resOrdered_plot$padj, col="blue", pch=18, cex=0.5)
qsig <-rownames(subset(resOrdered, padj < 0.05))
qsig_resOrdered_plot=resOrdered[qsig,]
points(qsig_resOrdered_plot$log2FoldChange, qsig_resOrdered_plot$padj, col="red", pch=8, cex=0.03)
abline(v=c(1,-1), col="black", lwd=1)
abline(h=0.05, col="red",lwd=1)
legend_text = c("<-1.5 or >1.5 Log2-Fold-Change", "p-value<0.05", "Adj-P-Value<0.05")
legend_text2 = c("P-Adj <0.05", "Log2-Fold-Change -1.5 to 1.5")
legend("topleft", legend_text,bty="n",pch = c(19,18,8), col=c("green","blue","red"))
legend("topright", legend_text2, bty = "n", lty=c(1, 1), col=c("red", "black"))
dev.off()
```
# Looks bad, lets remove the 3 outliers. ENSG00000204392, ENSG00000284934, ENSG00000285395
```{r}
resOrdered <- resOrdered[order(resOrdered$log2FoldChange),]
resOrdered <- resOrdered[-c(1:3),]
pdf(file="~/Results/A549_v_Ctrl_volcano(2).pdf")
plot(resOrdered$log2FoldChange, resOrdered$padj, col="black", pch=1, cex=1.1,  main="A549 vs Ctrl Volcano Plot")
fsig <-rownames(subset(resOrdered, log2FoldChange > 1.5 | log2FoldChange < -1.5,))
fsig_resOrdered_plot=resOrdered[fsig,]
points(fsig_resOrdered_plot$log2FoldChange, fsig_resOrdered_plot$padj, col="green", pch=19, cex=1.1)
psig <-rownames(subset(resOrdered, pvalue < 0.05))
psig_resOrdered_plot=resOrdered[psig,]
points(psig_resOrdered_plot$log2FoldChange, psig_resOrdered_plot$padj, col="blue", pch=18, cex=0.5)
qsig <-rownames(subset(resOrdered, padj < 0.05))
qsig_resOrdered_plot=resOrdered[qsig,]
points(qsig_resOrdered_plot$log2FoldChange, qsig_resOrdered_plot$padj, col="red", pch=8, cex=0.03)
abline(v=c(1,-1), col="black", lwd=1.5)
abline(h=0.05, col="red",lwd=1)
legend_text = c("<-1.5 or >1.5 Log2-Fold-Change", "p-value<0.05", "Adj-P-Value<0.05")
legend_text2 = c("P-Adj <0.05", "Log2-Fold-Change -1.5 to 1.5")
legend("topleft", legend_text,bty="n",pch = c(19,18,8), col=c("green","blue","red"))
legend("topright", legend_text2, bty = "n", lty=c(1.5, 1.5), col=c("red", "black"))
dev.off()
```
# A function that outputs:
# Up regulated genes + info 
# Down Regulated genes + info
# Perturbed genes as a character vector
```{r}
setwd("~/Results")
generate_output <- function(x, prefix){
  up_key <-  intersect(rownames(x)[which(x$log2FoldChange>=1.5)],
                       rownames(x)[which(x$padj<=0.05)])
  
  up_df <- as.data.frame((x)[which(rownames(x) %in% up_key),])
  up_df <- tibble::rownames_to_column(up_df, "ensembl_gene_id")
  
  info <- getBM(attributes=c("ensembl_gene_id",
                        "external_gene_name",
                        "chromosome_name",
                        "start_position",
                        "end_position",
                        "description"),
                        filters = c("ensembl_gene_id"),
                        values = up_df$ensembl_gene_id,
                        mart = mart)
  
  tmp <- merge(info, up_df, by="ensembl_gene_id")
  tmp <- tmp[order(-tmp$log2FoldChange),]
  write.table(tmp, paste0(prefix,"_up_reg.txt"),
              sep="\t", row.names = F, quote=F)
  
  down_key <- intersect(rownames(x)[which(x$log2FoldChange<=-1.5)],
                        rownames(x)[which(x$padj<=0.05)])
  
  down_df <- as.data.frame((x)[which(rownames(x) %in% down_key),])
  down_df <- tibble::rownames_to_column(down_df, "ensembl_gene_id")
  
  info <- getBM(attributes=c("ensembl_gene_id",
                        "external_gene_name",
                        "chromosome_name",
                        "start_position",
                        "end_position",
                        "description"),
                         filters = c("ensembl_gene_id"),
                         values = down_df$ensembl_gene_id,
                         mart = mart)
  
  tmp <- merge(info, down_df, by="ensembl_gene_id")
  tmp <- tmp[order(tmp$log2FoldChange),]
  write.table(tmp, paste0(prefix,"_down_reg.txt"), sep="\t", row.names = F, quote=F)
  
  sig_genes <- c(up_key, down_key)
  write.table(sig_genes, paste0(prefix,"_sig_list.txt"), 
              row.names = F, col.names = F, quote=F)
}
```
# Run the function for your DESeq results objects
```{R}
setwd("~/Results")
generate_output(lung_v_ctrl, "lung_v_ctrl")
#generate_output(melanoma_v_ctrl, "melanoma_v_ctrl")
#generate_output(lung_v_melanoma, "lung_v_melanoma")
```
# We will focus on Lung vs. Control for further visualisation and analysis. 
# Feel free to go back and try this out for the other comparisons made
```{r}
rm(list=ls())
#read in the DE file:
de_gene <- read.table("~/Results/lung_v_ctrl_sig_list.txt", sep = "\t", header=F)
counts <- read.table("~/Results/norm_counts.txt", header=T, sep="\t", row.names = "ensembl_gene_id")
log_counts <- log2(counts + 1)
key <- as.character(de_gene[,1])
mat <- as.data.frame((log_counts)[which(rownames(log_counts) %in% key),])
mat <- mat  %>% dplyr::select(-c(A375_1, A375_2, A375_3)) %>% slice(1:500)
pdf("~/Plots/lung_v_control_heatmap.pdf", height = 10, width=7.5)
pheatmap(mat, color=greenred(75), cluster_rows = T,
         fontsize=9, 
         show_rownames = F, scale = "row",
         cellheight = 1, cellwidth = 30) #scale really important here! :)
dev.off()
```
# GSEA
# Use the MSigDB GMT files to see which pathways our DE genes belong to.
```{R}
lung_v_ctrl_res <- read.csv("~/Results/lung_v_ctrl_up_reg.txt", sep = "\t")
fgsea_rank <- lung_v_ctrl_res %>%
  dplyr::select(external_gene_name, stat) %>%
  na.omit() %>%
  distinct() %>%
  group_by(external_gene_name) %>%
  summarize(stat=mean(stat))
rank <- deframe(fgsea_rank)
pathways.kegg <- gmtPathways("~/c2.cp.kegg.v7.0.symbols.gmt")
fgsea_kegg <- fgseaMultilevel(pathways=pathways.kegg, stats=rank) %>%
  as_tibble() %>%
  arrange(padj)
```
# This block will plot enrichment plots for every pathway that is statistically significant @ pval < 0.05. 
```{r}
setwd("~/Plots/fgsea_kegg")
filtered_pathway <- subset(fgsea_kegg, pval < 0.05)
filtered_pathway <- as.vector(filtered_pathway$pathway)
for (i in filtered_pathway){
    pdf(paste0(i,".pdf"),height=5,width=5) # change height and width parameter
    plt <- plotEnrichment(pathway = pathways.kegg[[i]], 
    gseaParam = 1, ticksSize = 0.5, stats= rank) + 
    labs(title=i) + theme(plot.title = element_text(hjust = 0.5, face="bold"))
    print(plt)
    dev.off()
}
```
# For demonstration, we will take the top 5 pathways from our fgsea result. 
# The code below converts the lists of genes into a binary matrix, and pulls the corresponding Log2FC information for each gene. 
# We can then plot important genes and which pathways they share, ordered in ascending Log2FC 
# will become apparent during tutorial (i hope)
```{R}
library(GOplot)
subset_pathway <- fgsea_kegg[1:10,]
list = list()
cols = c()
for (i in 1:nrow(subset_pathway)){
  row <- subset_pathway[i,]
  pathway <- row$pathway
  cols <- c(cols, pathway)
  list[[pathway]] = row$leadingEdge
}
binary_matrix <- list %>% lapply(table) %>% lapply(as.list) %>% 
    lapply(data.frame) %>% rbind_all()
binary_matrix <- t(binary_matrix)
colnames(binary_matrix) <- cols
binary_matrix[is.na(binary_matrix)] <- 0
LFC_key <- lung_v_ctrl_res[lung_v_ctrl_res$external_gene_name %in% rownames(binary_matrix),]
LFC_key <- LFC_key %>% dplyr::select(external_gene_name, log2FoldChange)
rownames(LFC_key) = LFC_key$external_gene_name
chord<- cbind(binary_matrix, LFC_key[rownames(binary_matrix), 2])
column_key <- ncol(chord)
chord <- chord[order(chord[,column_key]),]
colnames(chord)[column_key] <- "logFC"
pdf(file="~/Plots/Circos_plot.pdf", height =25,  width=25)
GOChord(chord, space = 0.02, gene.order = 'logFC', gene.size = 5)
dev.off()
```
